---
title: "Quantitative Genomics and Genetics 2020"
author: "Scott Kulm"
date: "2/13/2020"
output:
  html_document: default
  pdf_document: default
subtitle: Computer Lab 3
---

* Subsetting
* Plotting
* Time to debug

--------------------------------------------------------------------------------------

### 0. Announcements

General Homework Guidelines:

- Following Rachel's post on Piazza, all homework must be typed.  We recommend LaTeX if R is not involved, but you can also use Rmd or Microsoft Word if need be.

When submitting R code: 

- Use Rmarkdown and all working directories should be set to "./" if there are input datasets

- Submit your .Rmd script and compiled pdf/html. There will be penalties for scripts that fail to compile. (except errors related to missing dependencies - e.g. packages not installed on my computer, just make sure to mention which packages have been called).

Logistics:

- If you are having problems with CMS (including not being enrolled), please see me

--------------------------------------------------------------------------------------

## 1. Subsetting

### Quick Review

- Subsetting can be defined as accessing a specific element within a larger data structure

- This may sounds fancy, but we have already seen several examples, the most basic of which involves subsetting a vector using a single index within brackets

```{r}
x <- seq(5, 20, length.out = 10)
x[3]
```

- When you have multiple dimensions (matricies, data frames), you must specify the row or column

- If you leave a row or column spot blank then the entire row or column will be printed out

```{r}
x <- matrix(1:9, nrow = 3)
x[2,2]

x[2,]

x[,2]
```

- When using data frames you may use the dollar sign to specify columns

- Which is good motivation to always have columns with good column names

```{r}
x <- data.frame("numbers" = 1:10, "letters" = letters[1:10])
x$numbers
```

- Subsetting gets more complicated when we know the type, or quality of the subset elements, but we do not exactly know the indices of those elements

### A New Look at Subsetting

- A literal answer to this problem is using the which function

```{r}
which(x$numbers > 5)

which(x$letters == "a")
```

- "Under the hood", what's going on is R is making a comparison, generating a vector of Trues and Falses, and then finding and returning the index of all the Trues

- Note there are also some nice built in which functions that don't follow this idea

```{r}
which.max(x$numbers)
which.min(x$numbers)
```

- To get the actual values out of the data frame (or matrix, vector, ...), we do not need the actual index, we can also directly use the True and False values

- You can think that R is going element by element down your vector and asking "should this element be subset", it then looks at the matching element in the True and False vector for an answer

```{r}
true_and_false <- x$numbers > 5
for(i in 1:length(x$numbers)){
  if(true_and_false[i] == TRUE){
    print(x$numbers[i])
  }
}

#OR

x$numbers[x$numbers > 5]
```

**Question 1**

- What is the output of?

- Can you describe what's going on in a similar fashion as the above for loop

```{r, eval = FALSE}
letters[1:3 == 3]
```



- This process can get more complicated

```{r}
iris_subset <- iris[iris$Species == "setosa" & iris$Sepal.Length > iris$Sepal.Width[order(iris$Sepal.Width)][10] * 2, 1:3]
head(iris_subset)
```

- But can always be broken down again

```{r, eval = FALSE}
iris$Sepal.Width
# order(iris$Sepal.Width)
# iris$Sepal.Width[order(iris$Sepal.Width)]
# iris$Sepal.Width[order(iris$Sepal.Width)][10]
# iris$Sepal.Width[order(iris$Sepal.Width)][10] * 2
# iris$Sepal.Length > iris$Sepal.Width[order(iris$Sepal.Width)][10] * 2

# iris$Species
# iris$Species == "setosa"

# iris$Species == "setosa" & iris$Sepal.Length > iris$Sepal.Width[order(iris$Sepal.Width)][10] * 2
```

- A final note is the %in% keyword that allows you to make multiple or stype comparisons

```{r}
x <- 1:10
x[x %in% 4:7]

#Same as

x[x == 4 | x == 5 | x == 6 | x == 7]
```

- Just be sure to have you comparison the right way

```{r}
x[4:7 %in% x]
```

## 2. Visualization 

### plot() 

- plot() generates a X-Y plot

- The basic syntax is plot(x-axis, y-axis)

- You can add more information to the plot with axis labels and titles.

```{r ,fig.width = 4,fig.height = 4,fig.align='center' }
x <- seq(-5, 5, by = 0.5) 
y <- x - 3
plot(x, y, 
     xlab = "X axis label here", 
     ylab = "Y axis label here", 
     main = "Plot Title here")
```

- We just used base R plotting

- base R is good for simple plots, but becomes complicated and ugly for any more advanced examples


### Ggplot

- the popular alternative to base R is ggplot2 (we talked about this last week)

- ggplot2 is based on a grammer of graphics that makes plotting very modular

- The basic outline of making a ggplot is as follows:

```{r, eval = FALSE}
library(ggplot2)

ggplot(`dataframe you want to plot`, aes(x = `column name to make the x axis`, y = `column name to make the y axis`)) +
  geom_`whatever type of plot you want to make` + `you can simply add on in this way more modifications`
```


- let's try to remake the sample plot as before in ggplot


```{r ,fig.width = 4,fig.height = 4,fig.align='center' }
library(ggplot2)

x <- seq(-5, 5, by = 0.5) 
y <- x - 3
df <- data.frame(x,y)
ggplot(df,aes(x,y))+geom_point()
```

- How can I found out how to include axis labels?

- An amazing reference is https://ggplot2.tidyverse.org

- A broad overview on everything ggplot2 can do is https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

- Plotting is a great means for displaying and investigating distributions


```{r ,fig.width = 4,fig.height = 4,fig.align='center' }
df <- data.frame(x = rnorm(1000))
ggplot(df,aes(x))+geom_histogram()+
  labs(x = "quantile",y = "count",title = "Our normal distribution")
```

- looks like a normal distribution!

- rnorm produces 1000 random values drawn from a normal distribution

- this is akin to measuring the heights of 1000 people with mean height 0 and sd height 1

- R brings up an important point, the binwidth greatly changes how a histogram will look

```{r ,fig.width = 4,fig.height = 4,fig.align='center' }
ggplot(df,aes(x))+geom_histogram(bins=3)+
  labs(x="quantile",y="count",title="3 Bins")

ggplot(df,aes(x))+geom_histogram(bins=5)+
  labs(x="quantile",y="count",title="5 Bins")

ggplot(df,aes(x))+geom_histogram(bins=300)+
  labs(x="quantile",y="count",title="300 Bins")
```

- There are other ways to display the data

- We can convert the count to a density, which normalizes the counts

```{r ,fig.width = 4,fig.height = 4,fig.align='center' }
x <- rnorm(1000, mean = 0,sd = 1)
ggplot(df,aes(x)) + geom_density()
```

**Question 2**
How can I make the density plot above filled in with purple

```{r}
#purple plot instructions here
```

- What if we want to include both plots on the same screen, easy!

- Just add both geom's

```{r ,fig.width = 4,fig.height = 4,fig.align='center' }
ggplot(df,aes(x))+
  geom_histogram(aes(y=stat(density)),bins=30)+
  geom_density(color="red")+
  labs(x="quantile",y="density",title="Double Plots")
```

- Pulling random points from a distribution isn't the only, or likely best way to investigate a distribution

- Each distribution has functions called r(random), d(density), p(cumulative), q(quantile)

- Let's start with density as we have already been looking at it

```{r}
dnorm(0,mean = 0,sd = 1)
dnorm(2,mean = 0,sd = 1)
dnorm(-1,mean = 0,sd = 1)
```

- Let's plot it!

```{r ,fig.width = 4,fig.height = 4,fig.align='center' }
df <- data.frame(x=seq(-3,3,0.1),d=dnorm(seq(-3,3,0.1)))
ggplot(df,aes(x,d))+geom_point()
```

- Next, let's take a look at the cumulative distribution

```{r}
pnorm(0,mean = 0,sd = 1)
pnorm(2,mean = 0,sd = 1)
pnorm(-1,mean = 0,sd = 1)
```

- Let's plot this too!

```{r ,fig.width = 4,fig.height = 4,fig.align='center' }
df$p <- pnorm(df$x)
ggplot(df) + geom_point(aes(x,d), color="red") + geom_point(aes(x,p), color="blue")
```

- The cumulative distribution logically reaches 0.5 just above the peak of the density

**Question 3**
How can I include a legend with this plot? (Hint you may need just one geom)
```{r}


```


- Lastly we have q, which converts a cumulative density back into the quantile it came from

```{r}
qnorm(0.5,mean = 0,sd = 1)
qnorm(pnorm(0.5),mean = 0,sd = 1)
```

- The "rules" for d,c,r,p also work for every other distribution, although the parameters will change

- For example there is the binomial distribution, which indicates the number of successes in a number of trials

```{r}
rbinom(n=1, size = 10, prob = 0.3)
dbinom(4, size=10, prob=0.5)
getwd()
```

- Let's make a plot comparing the densities of the binomial and normal distributions

```{r ,fig.width = 4,fig.height = 4,fig.align='center' }
normVals <- dnorm(seq(-3,3,length.out = 100),mean=0,sd=1)
binomVals <- dbinom(1:100,size=100,prob=0.5)

#standard the values to min of 0 and max of 1
normVals <- (normVals-min(normVals))/(max(normVals)-min(normVals))
binomVals <- (binomVals-min(binomVals))/(max(binomVals)-min(binomVals))

df <- data.frame(xNorm=seq(-3,3,length.out = 100),normVals,xBinom=1:100,binomVals)

ggplot(df) + geom_line(aes(xNorm,normVals),color="red") +
  geom_point(aes(xNorm,binomVals),color="blue")+
  labs(x="X-Value",y="Density")
```

- While the two distributions are similar, they are clearly not the same



### Exercise
The midwest dataframe has some great data on various midwest counties.  Use the dataframe, fill in the pseudo-code below, and ultimately make the 3 following plots

```{r}
#first plot - compare poptotal and area
```

```{r}
library(ggplot2)
library(cowplot)
theme_set(theme_cowplot())

ggplot(midwest, aes(area, poptotal)) + geom_point()
```

```{r}
#interesting, there seems to be one clear outlier, let's find out it's name and label it
  #1 - get the name of the county with population over 4e+6
  #2 - create a new column of the data frame called "plot_labels", every element will be NA
  #3 - change the value of plot_labels at the index of the county determined in #1 to the name of the county
  #4 - convert the inmetro column to a factor data type
  #5 - create the below plot, possibly using geom_label (https://ggplot2.tidyverse.org/reference/geom_text.html),
  #    and color = ..., and labs()
  
#while not included here for simplicity, you add on geom_smooth to see a linear interpolation
```

```{r}
big_county <- midwest$county[midwest$poptotal > 4e+6]
midwest_plot <- data.frame(midwest)
midwest_plot$plot_labels <- NA
midwest_plot$plot_labels[midwest$poptotal > 4e+6] <- big_county
midwest_plot$inmetro <- as.factor(midwest$inmetro)

ggplot(midwest_plot, aes(area, poptotal, color = inmetro, label = plot_labels)) + geom_point() + geom_text() +
  labs(x = "Area", y = "Population Total", color = "In Metro Area")
```


```{r}
#let's switch over to boxplots
  #1 - subset the midwest data frame, removing labeled county above and all counties with population less than the 10th percentile of total population
  #2 - create a new column that indicates whether perccollege is over 20
  #3 - finally, make the plot below
```

```{r}
midwest_plot <- data.frame(midwest)
midwest_plot <- midwest_plot[!(midwest$county == big_county | midwest$poptotal < quantile(midwest$poptotal, 0.1)),]
midwest_plot$college <- 0
midwest_plot$college[midwest_plot$percollege > 20] <- 1
midwest_plot$college <- as.factor(midwest_plot$college)

ggplot(midwest_plot, aes(state, poptotal, fill = college)) + geom_boxplot() + 
  labs(x = "State", y = "Total Population", fill = "% College > 20") +
  scale_fill_discrete(labels = c("no", "yes"))


```